---
title: "Class 11: Structural Bioinformatics I"
author: "Serina Huang"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PDB database composition statistics

```{r}
pdbstats <- read.csv("Data Export Summary.csv", row.names = 1)
```

Let's look at the table:
```{r}
library(knitr)
kable(pdbstats)
```

Q1. Determine percentage of structures solved by X-ray and electron microscopy. What proportion of structures are protein?
```{r}
# total number of entries
tot = sum(pdbstats$Total)
percent <- round(pdbstats$Total / tot * 100, 2)
```

A1. There are `r percent[1]`% X-ray structures and `r percent[3]`% EM structures in the PDB as of `r Sys.time()`.
```{r}
nstats <- pdbstats
nstats$Percent <- percent
kable(nstats)
```

The proportion of protein entries:
```{r}
percent_protein <- round( sum((nstats$Proteins)) / tot * 100, 2 )
```

There are `r percent_protein`% of proteins in the PDB.

## Introduction to Bio3D
```{r}
library(bio3d)
pdb <- read.pdb("1HSG")
pdb
attributes(pdb)
```

Let's plot the B factor against each residue.
```{r}
plot.bio3d(pdb$atom$b[pdb$calpha], sse = pdb, typ = "l", ylab = "B-factor")
```

Q6. How many amino acid residues are there in this pdb object? What are the two non-protein residues?

A6. There are 1686 total atoms. The two non-protein residues are 127 water compounds and 1 non-protein ligand, i.e. Merck's inhibitor.

Atom selection:
```{r}
ca.inds <- atom.select(pdb, "calpha")
ca.inds
# check out atoms in the alpha carbon chain
head(pdb$atom[ca.inds$atom,])
```

```{r}
ligand.pdb <- atom.select(pdb, "ligand", value = TRUE)
write.pdb(ligand.pdb, file = "MK1.pdb")
```

## Aligning multiple structures
```{r}
ids <- c("1TND_B","1AGR_A","1TAG_A","1GG2_A","1KJY_A","4G5Q_A")
files <- get.pdb(ids, split = TRUE)

# extract and align the chains of interest
pdbs <- pdbaln(files, fit = TRUE)
# print a summary of pdbs object
pdbs
```

```{r}
# find invariant (most rigid) core
core <- core.find(pdbs)
# superimpose all structures to core
pdbs$xyz <- pdbfit(pdbs, core)
# perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```

```{r}
# visualize the first PC
pc1 <- mktrj(pc.xray, pc = 1, file = "pc_1.pdb")
# load on VMD, index coloring method, tube drawing method, all selected atoms
```


