---
title: "Lab 8"
author: "Serina Huang"
date: "October 25, 2018"
output: github_document
---

## Section 1. PCA of UK food data
```{r}
x <- read.csv("https://bioboot.github.io/bimm143_F18/class-material/UK_foods.csv")
dim(x)
```

Let's check out the data:
```{r}
head(x)
```

Fix the data so each country gets one column:
```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
dim(x)
```

Generate a table:
```{r}
knitr::kable(x, caption = "The full UK foods data table")
```

Can you extract any useful information from this table? I thought not. Let's try representing it in a heatmap.
```{r}
heatmap(as.matrix(x))
```

What are the relationships between the variables (types of food)? You can't do it without PCA. Now, we need to transpose our matrix because the `prcomp()` function takes observations in rows and variables in columns.
```{r}
pca <- prcomp( t(x) )
summary(pca)
```

Which PC has the maximum variance? The summary shows that PC1 captures 67% of variance and PC2 captures 29%, for a total of 96%. Usually, 70% is good 'nuff.
```{r}
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2", xlim = c(-270, 500))

text(pca$x[,1], pca$x[,2], colnames(x))
```

There are two ways to find the variance of each PC. First, we can calculate it ourselves by squaring the standard deviation:
```{r}
v <- round( pca$sdev^2 / sum(pca$sdev^2) * 100 )
v
```

Second, we can use the `summary` as we've done previously.
```{r}
z <- summary(pca)
z$importance
```

Represent the summary in a scree plot:
```{r}
barplot(v, xlab = "Principal Component", ylab = "Percent Variation")
```

We can also find the cumulative sum of the variance.
```{r}
cumsum(v)
```

Let's look at the loading score of each observation in PC1.
```{r}
# Perpendicular labels
barplot( pca$rotation[,1], las = 2)
```

How to interpret: fresh potatoes (most positive) is the observation that pushes N. Ireland to the right side of the plot. Fresh fruit (most negative) is the observation that pushes Wales, England, and Scotland to the left side of the plot.

We can see the same information with biplot.
```{r}
biplot(pca)
```

## Section 2. PCA of example RNA-seq data
```{r}
data.matrix <- matrix(nrow=100, ncol=10)
# Label rows with gene names
rownames(data.matrix) <- paste0("gene", 1:100)
# Label columns with sample cells
colnames(data.matrix) <- c(
  paste0("wt", 1:5),
  paste0("ko", 1:5)
)

head(data.matrix)

# Fill in fake read counts
for (i in 1:100) {
  wt.values <- rpois(5, lambda = sample(x = 10:1000, size = 1))
  ko.values <- rpois(5, lambda = sample(x = 10:1000, size = 1))
  
  data.matrix[i,] <- c(wt.values, ko.values)
}

head(data.matrix)
```

It's PCA time!
```{r}
pca <- prcomp( t(data.matrix), scale = TRUE)
plot(pca$x[,1], pca$x[,2])

# Variance captured per PC
pca.var <- pca$sdev^2
# Convert variance into percent variance (1 decimal point)
pca.var.per <- round(pca.var / sum(pca.var) * 100, 1)
pca.var.per
# Use this to generate scree plot
barplot(pca.var.per, main = "Scree Plot", xlab = "Principal Component", ylab = "Percent Variation")
```

Color code our PCA plot:
```{r}
# Make a color vector for wt and ko samples
colvec <- colnames(data.matrix)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col = colvec, pch = 16,
     xlab = paste0("PC1 (", pca.var.per[1], "%)"),
     ylab = paste0("PC2 (", pca.var.per[2], "%)"))

# Label samples to the right of the first five points, and to the left of the last five points
text( pca$x[,1], pca$x[,2], labels = colnames(data.matrix), pos = c(rep(4,5), rep(2,5)) )
```

Let's look at the loading scores of each gene (observation) in PC1.
```{r}
loading_scores <- pca$rotation[,1]
# Since we're interested in effect magnitude, use absolute value of loading scores. Then rank them.
gene_scores <- abs(loading_scores)
gene_score_ranked <- sort(gene_scores, decreasing = TRUE)

# Get the name of the top 10 genes
top_10_genes <- names(gene_score_ranked[1:10])
top_10_genes
```

## Section 3. PCA of protein structure data