---
title: "Class 13: Structural Bioinformatics II"
author: "Serina Huang"
date: "November 13, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1. _In silico_ docking of drugs to HIV-1 protease
```{r}
library(bio3d)
file.name <- get.pdb("1HSG")
hiv <- read.pdb(file.name)
```

##### Q1. What are the 2 non-protein **resid** values in this structure? What does **resid** correspond to and how would you get a listing of all residue values in this structure?
```{r}
# Quick summary of contents of this pdb structure
hiv
```

##### A1. HOH, water, and MK1, a Merck drug compound.

### Split into separate protein and ligand files
```{r}
# The protein-only portion is our receptor
prot <- trim.pdb(hiv, "protein")
prot
# Write this into a file
write.pdb(prot, file = "1hsg_protein.pdb")
# Inspect the resulting file that all are proteins
```

Do the same for ligand:
```{r}
lig <- trim.pdb(hiv, "ligand")
lig
# The only non-protein residue is MK1, and no HOH
write.pdb(lig, file = "1hsg_ligand.pdb")
# Inspect again
```

Note: if there is more than 1 ligand in a structure, you can get the ligand at a certain residue, e.g. at 902.
