---
title: "hw06"
author: "Serina Huang"
date: "10/22/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The protein_drug( ) function

##### Function:
The protein_drug( ) function trims the protein data, keeping residues with an alpha chain and the CA element. It then plots the temperature factor along each amino acid residue in a line scatter plot, with an optional secondary structure estimation in the margins. To turn off margins, set margins = NULL in the function argument.

##### Input:
The four letter PDB identifier of the protein.

##### Output:
None.

***

First, install the bio3d library:
```{r install library}
library(bio3d)
```

The entire code for the function is produced below:
```{r content of function}
protein_drug <- function(x, margins = TRUE) {
  
  # load data of user specified protein
  protein <- read.pdb(x)
  
  # only keep amino acids in the alpha chain and of the CA element type
  chainA <- trim.pdb(protein, chain = "A", elety = "CA")
  
  # among amino acids chosen, select the column for temperature factor
  temp_fact <- chainA$atom$b
  
  # scatter plot temperature factor and secondary structure estimation
  # in the margins as default
  if (margins == TRUE) {
    plotb3(temp_fact, sse = chainA, type = "l", ylab = "Temperature Factor")
  }
  else {
    plotb3(temp_fact, sse = NULL, type = "l", ylab = "Temperature Factor")
  }
  
}
```

##### Example:
```{r example}
x = "1E4Y"
protein_drug(x)
```